Containers and R---the Rockerverse and beyond

### 2020-01-13 manuscript.tex
update tex file

% !TeX root = RJwrapper.tex
\title{Containers and R---the Rockerverse and beyond}
\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Ben Marwick}

@@

Docker, an application and service provide by the eponymous company, has
in just a few short years risen to prominence for development, testing,
deployment and distribution of computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
related approaches such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerization''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operationg systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualization. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, an important
scalability opportunity is being provided. But Docker improved on
virtualization by accessing the host system---generally Linux---through
a much thinner and smaller shim than a full operating system emulation
or virtualization. This makes for more efficient use of system resources
\citep{felter_updated_2015} and allowed another order of magnitude in
terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
Docker makes use of Linux kernel features, its importance was large
enough so that some required aspects of running Docker have been added
to other operating systems to support Docker there more efficiently too
\citep{microsoft_linux_2019}. The success even lead to standardisation
and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker (as an ``application'') is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive, and novel, proposition which has lead to widespread
adoption and use of Docker in a variety of domains, \emph{e.g.}, cloud
computing infrastructure \citep{Bernstein2014}, data science
\citep{boettiger_introduction_2015}, and edge computing
\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
a natural match for ``cloud deployment'' which runs (or at least appears
to run) ``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

@@

\textbf{FastR} is ``A high-performance implementation of the R
programming language, built on GraalVM''
(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
also claims superior performance but also targets full compatibility
with base R
\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but \texttt{nuest/fastr}
images and \texttt{Dockerfile} are independently maintained on
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
respectively.

@@

\href{http://bioconductor.org/}{\emph{Bioconductor}} is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \_Bioconductor\_has a semiannual release
cycle, each release is associated with a particular version of R. Docker
images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor} `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
installs packages based on the R version, and therefore uses
\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
\emph{Bioconductor} selects the desired version of R from Rocker, adds
the BiocManager CRAN package for installing appropriate versions of
\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
image with an informative tag (R\_version\_Bioc\_version). The images
are summarized on the \emph{Bioconductor} web site
(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
(\url{https://github.com/Bioconductor/bioc_docker}), and available to
the community through \href{https://hub.docker.com/}{DockerHub}. Past
and current combinations of R and \emph{Bioconductor} are therefore
accessible via a specific docker tag.

@@

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\ref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user. For current Windows Server (2016 and later)
and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{\href{https://github.com/nuest/rocker-win}{\texttt{rocker-win}}}
is a proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. New R versions are only added on demand. The
explorative project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable actually,
as R ships pre-compiled binaries for Windows.

@@

\label{rhub} The \textbf{R-hub} project provides \_``a collection of
services to help R package development'', with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \pkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on Docker
Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
\texttt{see}rhub::local\_check\_linux\_images()\texttt{for\ a\ full\ list),\ and\ are\ intended\ for\ debugging\ (see\ Section\&nbsp;\textbackslash{}ref\{development\})\ not\ for\ reproducible\ workflows.\ \_The\ rationale\_\ of\ not\ re-using\ the\ Rocker\ image\ stack\ is\ the\ high\ number\ of\ operatings\ systems\ and\ configurations,\ which\ can\ be\ more\ uniformly\ maintained\ in\ a\ seperate\ suite\ of\ images.\ Also,\ a\ common\ structure\ across\ the\ respective\ image\ stacks\ of\ the\ operating\ systems,\ e.g.,\ for}debian\texttt{,\ adding\ GCC\ in}debian-gcc\texttt{and\ then\ R\textquotesingle{}s\ development\ version\ in}debian-gcc-devel`.
The platform also has a specific user configuration differing from the
setup covering typical applications of Rocker images, and the images
include considerably more software, such as a fairly complete LaTeX
installation, to provide an experience closer to CRAN.

There are also independent projects installing R or RStudio on
\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
beyond the proof of concept
stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
Alpine images are often used for minimalistic and thereby small and
secure containers, e.g., in hardware with limited storage, though this
advantage quickly deteriorates if more features are needed, e.g.~if
installing RStudio, package building envrionments, or powerful data
science tools. Furthermore the distribution uses the
\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
experiences
exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
\href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr}
only has a single and quite short \texttt{Dockerfile}, showing the easy
installation of the latest R from Alpine sources. These images are used
\ldots{}

\href{https://github.com/ismailsunni/altRnative/}{\texttt{altRnative}}
is an experimental R package for running the same code across multiple
containerised versions of R, intended for comparison across operating
systems and different implementations of R. It comes with a collection
of \texttt{Dockerfile}s and corresponding images
\footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.

@@

The notion of \emph{Data Science} is widely discussed among all academic
disciplines. Using computers and complex software to make sense of
(large amounts of) collected or simulated data is a cross-cutting
phenomenon leading to \emph{X Data Science} disciplines, seminars, and
study majors, there are also Docker images with commonly used tools and
software stacks pre-installed that are often used in combination for
analysing data. A large number of bespoke computing environments are
created this way. Here we only mention a few image stacks as examples
that are not based on Rocker images but prominently include
installations of R.

@@

Interfacing with the Docker daemon is typically done through the Docker
Command Line Interface (CLI). However, moving back and forth between an
R console and the command line can create friction in workflows and
reduce reproducibility. A number of first-order R packages provide a
interface to the Docker CLI, allowing to automate interaction with the
Docker CLI from an R console.

Each of these packages has particular advantages as they provide
function wrappers for interacting with the Docker CLI at different
stages of a container's life cycle. Examples of such interactions are
installing the Docker software, creating \texttt{Dockerfile}s, building
an image, and launching a container on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the
use-case at hand.

@@

\textbf{stevedore} (\url{https://cran.r-project.org/package=stevedore})

\textbf{dockyard} (\url{https://github.com/thebioengineer/dockyard}) has
the goal of lowering barrier to creating \texttt{Dockerfile}s and Docker
images, and deploying Docker containers. The way docker images are made
is by reading and following a set of instructions found in
\texttt{Dockerfile}s. This requires stepping out of the R ecosystem and
into an infrastructure mindset where the user needs to specify the
environment they require. Because \texttt{Dockerfile}s can be read as a
series of steps, this lends itself well to the \texttt{tidyverse} style
of programming where each step is piped into the next. dockyard has a
\texttt{Dockerfile} object that either starts out as an empty
\texttt{Dockerfile}, or can use existing \texttt{Dockerfile} to use as a
template. This combined with a series of functions that are analogous to
the steps commonly found in a \texttt{Dockerfile}, it makes building
Docker images much more approachable to R users. There are also wrappers
for commonly used steps for R programmers like adding libraries and
copying files into the R Docker containers. Once the \texttt{Dockerfile}
is created, dockyard has the functionality to build the image and deploy
it.

\textbf{dockermachine} (\url{https://github.com/cboettig/dockermachine})

\textbf{AzureContainers}
(\url{https://cran.r-project.org/package=AzureContainers}) is an
interface to a number of container-related services in Microsoft's Azure
cloud, namely Container Instances, Container Registry and Kubernetes
Service. While it is mainly intended for working with Azure, as a
convenience feature it includes lightweight, cross-platform shells to
Docker, kubectl and helm. These can be used to create and manage
arbitrary Docker images and containers, as well as Kubernetes clusters
on any platform or cloud service.

\textbf{babelwhale}
(\url{https://cran.r-project.org/package=babelwhale}) allows executing
and interacting with containers, which can use either Docker or
Singularity as a backend. provides a unified interface to interact with
Docker and Singularity containers. You can execute a command inside a
container, mount a volume or copy a file.

\textbf{harbor} (\url{https://github.com/wch/harbor}

@@

\begin{itemize}
\tightlist
\item
  \href{https://github.com/ColinFay/dockerfiler/}{\texttt{dockerfiler}}
\item
  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
\item
  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}
\item
  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
\item
  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
  with Docker}
\end{itemize}

\textbf{\href{https://nanx.me/liftr/}{\emph{liftr}}} \citep{liftr2019}
aims to solve the problem of persistent reproducible reporting in
statistical computing. Currently, the R Markdown format and its backend
compilation engine \emph{knitr} offer a \emph{de facto} standard for
creating dynamic documents \citep{xie2018}. However, the reproducibility
of such content authoring environments is often limited to individual
machines --- it is not easy to replicate the system environment
(libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative
document authoring and creating large-scale document building services.
\emph{liftr} solves this reproducibility problem by bringing Docker to
the game. In essence, \emph{liftr} helps R Markdown users create and
manage Docker containers for rendering the documents, thus make the
computations utterly reproducible across machines and systems.

@@

\hypertarget{deployment-to-cloud-services-div}{%
\subsection{Deployment to cloud services
{[}div{]}}\label{deployment-to-cloud-services-div}}

@@

\hypertarget{deployment-to-the-cloud-hongooi}{%
\subsection{\texorpdfstring{Deployment to the cloud
\citep{hongooi}}{Deployment to the cloud {[}@hongooi{]}}}\label{deployment-to-the-cloud-hongooi}}

>>>>

% !TeX root = RJwrapper.tex
\title{Containers and R---the Rockerverse and beyond}
\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Sean Lopp, Ben Marwick}

@@

Docker, an application and service provide by the eponymous company, has
in just a few short years risen to prominence for development, testing,
deployment and distribution of computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
related approaches such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerization''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operationg systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualization. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, an important
scalability opportunity is being provided. But Docker improved this
compartmentalization by accessing the host system---generally
Linux---through a much thinner and smaller shim than a full operating
system emulation or virtualization. This containerization is also called
operating-system-level virtualization
\ref{wikipedia_contributors_os-level_2020}. Typically a container runs
one process, whereas virtualization may run whole operating systems at a
larger footprint. This makes for more efficient use of system resources
\citep{felter_updated_2015} and allowed another order of magnitude in
terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
Docker makes use of Linux kernel features, its importance was large
enough so that some required aspects of running Docker have been added
to other operating systems to support Docker there more efficiently too
\citep{microsoft_linux_2019}. The success even lead to standardisation
and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition and Docker's very easy to use user
interface has lead to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure \citep{Bernstein2014},
data science \citep{boettiger_introduction_2015}, and edge computing
\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
a natural match for ``cloud deployment'' which runs, or at least appears
to run, ``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

@@

\textbf{FastR} is ``A high-performance implementation of the R
programming language, built on GraalVM''
(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
also claims superior performance but also targets full compatibility
with base
R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but \texttt{nuest/fastr}
images and \texttt{Dockerfile} are independently maintained on
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
respectively.

@@

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \_Bioconductor\_has a semiannual release
cycle, each release is associated with a particular version of R. Docker
images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor} `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
installs packages based on the R version, and therefore uses
\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
\emph{Bioconductor} selects the desired version of R from Rocker, adds
the BiocManager CRAN package for installing appropriate versions of
\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
image with an informative tag (R\_version\_Bioc\_version). The images
are summarized on the \emph{Bioconductor} web site
(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
(\url{https://github.com/Bioconductor/bioc_docker}), and available to
the community through \href{https://hub.docker.com/}{DockerHub}. Past
and current combinations of R and \emph{Bioconductor} are therefore
accessible via a specific docker tag.

@@

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\nameref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user. For current Windows Server (2016 and later)
and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. New R versions are only added on demand. The
explorative project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable actually,
as R ships pre-compiled binaries for Windows.

@@

\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
services to help R package development''}, with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \pkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on Docker
Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
are intended for debugging (see Section~\nameref{development}) not for
reproducible workflows. \emph{The rationale} of not re-using the Rocker
image stack is the high number of operatings systems and configurations,
which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the
operating systems, e.g., for \texttt{debian}, adding GCC in
\texttt{debian-gcc} and then R's development version in
\texttt{debian-gcc-devel}. The platform also has a specific user
configuration differing from the setup covering typical applications of
Rocker images, and the images include considerably more software, such
as a fairly complete LaTeX installation, to provide an experience closer
to CRAN.

\textbf{RStudio} also maintains a set of
\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
stacks include an \href{https://github.com/rstudio/r-builds}{opinionated
R installation} and ensure the R installation and profile is consistent
across the different Linux distributions. The images also install R in a
versioned directory and use a minimal set of build and runtime
dependencies, Furthermore these images are responsible for creating
pre-compiled binary R packages for Linux, which supplement the binaries
built by CRAN. These pre-compiled packages dramatically decrease the
installation time of R packages on Linux and subsequently can decrease
the build time of Docker images \citep{lopp_package_2019}.

There are also independent projects installing R or RStudio on
\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
beyond the proof of concept
stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic
and thereby small and secure containers, e.g., in hardware with limited
storage. R-hub's proof of concept
\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
to the extreme with compressed image size of 20MB and bespoke install
scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are
needed, e.g.~if adding compilers, packages, or especially if installing
powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the
\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
experiences
exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's `slim' variant and removing capabilities, one
could probalby achieve similarly small image sizes on a more common
stack.

\textbf{\pkg{altRnative}}
(\url{https://github.com/ismailsunni/altRnative/}) is an experimental R
package for running the same code across multiple containerised versions
of R, intended for comparison across operating systems and different
implementations of R. It comes with a collection of \texttt{Dockerfile}s
and corresponding images in multiple combinations, currently including,
e.g., MRO, FastR, Fedora, and
TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.

@@

\emph{Data Science} is a widely discussed topic among all academic
disciplines \citep[cf. e.g.,][]{donoho_50_2017}. Using computers and
complex software to make sense of (large amounts of) collected or
simulated data is a cross-cutting phenomenon leading to \emph{X Data
Science} disciplines, seminars, and study majors. The practice to apply
a combination tools and software stacks for analysing data and the
cross-cutting skillsets connected with data science puts the management
of computing environments and the challenges of reproducibility high up
on the agenda. This awareness leads to a large number of bespoke Docker
images but also some commonly used image stacks, ranging from base
images with very common tools to `catch all' images including \emph{a
lot} of software. Here we only mention a few image stacks as examples,
which are not based on Rocker images or stem from the R community, but
prominently include installations of R. A general description of what to
consider when creating a Docker image used for data science projects in
R is provided on
\href{https://environments.rstudio.com/docker}{environments.rstudio.com}.

@@

\label{interfaces}

Interfacing with the Docker daemon is typically done through the
\href{https://docs.docker.com/engine/reference/commandline/cli/}{Docker
Command Line Interface} (Docker CLI). However, moving back and forth
between an R console and the command line can create friction in
workflows and reduce reproducibility. A number of first-order R packages
provide a interface to the Docker CLI, allowing to automate interaction
with the Docker CLI from an R console.

Each of these packages has particular advantages as they provide
function wrappers for interacting with the Docker CLI at different
stages of a container's life cycle. Examples of such interactions are
installing the Docker software, creating \texttt{Dockerfile}s
(\pkg{dockerfiler}), \pkg{containerit}), building images and launching a
containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the
cloud. As such, the choice of which package is most useful depends on
the use-case at hand, but also the users level of expertise.

@@

\textbf{\pkg{stevedore}}
(\url{https://cran.r-project.org/package=stevedore})

\textbf{\pkg{AzureContainers}}
(\url{https://cran.r-project.org/package=AzureContainers}) is an
interface to a number of container-related services in Microsoft's
\href{https://azure.microsoft.com/}{Azure Cloud}, namely
\href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
Instances},
\href{https://azure.microsoft.com/en-us/services/container-registry/}{Container
Registry}, and
\href{https://azure.microsoft.com/en-us/services/kubernetes-service/}{Kubernetes
Service}. While it is mainly intended for working with Azure, as a
convenience feature it includes lightweight, cross-platform shells to
Docker and Kubernetes (tools \texttt{kubectl} and \texttt{helm}). These
can be used to create and manage arbitrary Docker images and containers,
as well as Kubernetes clusters on any platform or cloud service.

\textbf{\pkg{babelwhale}}
(\url{https://cran.r-project.org/package=babelwhale}) allows executing
and interacting with containers, which can use either Docker or
Singularity as a backend. The package provides a unified interface to
interact with Docker and Singularity containers. Users can, for example,
execute a command inside a container, mount a volume or copy a file.

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}

\textbf{\pkg{dockyard}}
(\url{https://github.com/thebioengineer/dockyard}) has the goal of
lowering barrier to creating \texttt{Dockerfile}s and Docker images, and
deploying Docker containers. The way docker images are made is by
reading and following a set of instructions found in
\texttt{Dockerfile}s. This requires stepping out of the R ecosystem and
into an infrastructure mindset where the user needs to specify the
environment they require. Because \texttt{Dockerfile}s can be read as a
series of steps, this lends itself well to the \texttt{tidyverse} style
of programming where each step is piped into the next. dockyard has a
\texttt{Dockerfile} object that either starts out as an empty
\texttt{Dockerfile}, or can use existing \texttt{Dockerfile} to use as a
template. This combined with a series of functions that are analogous to
the steps commonly found in a \texttt{Dockerfile}, it makes building
Docker images much more approachable to R users. There are also wrappers
for commonly used steps for R programmers like adding libraries and
copying files into the R Docker containers. Once the \texttt{Dockerfile}
is created, dockyard has built-in functionality to build the image and
run a container.

\textbf{\pkg{dockermachine}}
(\url{https://github.com/cboettig/dockermachine}) is an R package to
provide a convenient interface to
\href{https://docs.docker.com/machine/overview/}{Docker Machine} from R.
Docker Machine's CLI tool \texttt{docker-machine} allows users to create
and manage virtual host on local computers, local data centers, or at
cloud providers. A local Docker installation can be configured to
forward all commands issues on the local Docker CLI to a selected
(remote) virtual host fully transparent. For local Docker Machine was
especially crucial in early days of Docker, when no native Docker
support was available for Mac or Windows computers, but remains relevant
for provisioning on remote systems. The package has not received any
updates for two years, but was functional with a current version of
\texttt{docker-machine} (\texttt{0.16.2}) and potentially lowers the
barriers for R users to run containers on various hosts, if using the
Docker Machine CLI is perceived as a barrier.

@@

\begin{itemize}
\tightlist
\item
  \href{https://github.com/ColinFay/dockerfiler/}{\texttt{dockerfiler}}
\item
  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
\item
  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}, see also
  \url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
\item
  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
\item
  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
  with Docker}
\item
  \url{https://github.com/gigantum/base-images}
\end{itemize}

\textbf{\pkg{liftr}} (\url{https://nanx.me/liftr/}) \citep{liftr2019}
aims to solve the problem of persistent reproducible reporting in
statistical computing. Currently, the R Markdown format and its backend
compilation engine \emph{knitr} offer a \emph{de facto} standard for
creating dynamic documents \citep{xie2018}. However, the reproducibility
of such content authoring environments is often limited to individual
machines --- it is not easy to replicate the system environment
(libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative
document authoring and creating large-scale document building services.
\emph{liftr} solves this reproducibility problem by bringing Docker to
the game. In essence, \emph{liftr} helps R Markdown users create and
manage Docker containers for rendering the documents, thus make the
computations utterly reproducible across machines and systems.

@@

\hypertarget{deployment-to-cloud-services}{%
\subsection{Deployment to cloud
services}\label{deployment-to-cloud-services}}

@@

\hypertarget{deployment-to-the-cloud}{%
\subsection{Deployment to the cloud}\label{deployment-to-the-cloud}}
